---
title: "spatial_clustering"
output: html_document
date: "2025-07-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Spatial Clustering for UK 2022 GHG emission 


### 1.0 data cleaning-loading packages
```{r}
library(readr);    library(dplyr)
library(sf);       library(spdep)
library(FactoMineR);library(factoextra)
library(cluster);  library(tmap)
tmap_mode("plot")
```

### 1.1 reading GHG data for 2022
```{r}

ghg <- read_csv(
  "data/2005-2022-local-authority-ghg-emissions-csv-dataset.csv",
  show_col_types = FALSE
)

ghg22 <- ghg %>% filter(`Calendar Year` == 2022)
```

### 1.2 selecting out all sectors I need
```{r}
# select out all the sectors 
# Also I only need the GHG is CO2
library(dplyr)

target_sectors <- c(
  "Agriculture",
  "Industry",
  "Commercial",
  "Domestic",
  "LULUCF",
  "Public Sector",  
  "Transport",
  "Waste"
)

# filter out GHG == "CO2"
CO2 <- ghg22 %>%
  filter(
    `LA GHG Sector` %in% target_sectors,
    `Greenhouse gas` == "CO2"
  )

# summarise all CO2 territorial emission 
CO2_sum <- CO2 %>%
  group_by(`Local Authority Code`, `LA GHG Sector`) %>%
  summarise(kt_CO2 = sum(`Territorial emissions (kt CO2e)`), .groups = "drop")

# turn multi-variable into columns
library(tidyr)
CO2_sector <- CO2_sum %>%
  pivot_wider(names_from = `LA GHG Sector`,
              values_from = kt_CO2,
              values_fill  = 0)

CO2emission <- CO2_sector %>%
  left_join(select(ghg22, `Local Authority`,`Local Authority Code`, `Mid-year Population (thousands)`, `Area (km2)`) %>% distinct(),
            by = "Local Authority Code") %>%
  mutate(across(all_of(target_sectors),
                ~ .x / (`Mid-year Population (thousands)` * 1e3),  # t CO2 / person
                .names = "{.col}_pc"))

#clean the df
CO2emission <- CO2emission %>%
  select(
    `Local Authority Code`,
    `Local Authority`,
    `Mid-year Population (thousands)`,
    `Area (km2)`,
    Agriculture,
    Agriculture_pc,
    Industry,
    Industry_pc,
    Commercial,
    Commercial_pc,
    Domestic,
    Domestic_pc,
    `Public Sector`,
    `Public Sector_pc`,
    Transport,
    Transport_pc,
    Waste,
    Waste_pc,
    LULUCF,
    LULUCF_pc
  )

```

p.s. Here territorial emissions covers all emissions happened in that geographical area
The subset table is CO2 emissions within the scope of influence of LAs, which Only CO₂ sources that local councils can realistically shape through local policy and services.

###1.3 Loading UK LAD map and merge it with CO2 emission
```{r}
# 1. read map
UKlad <- sf::st_read("data/LAD_boundaries/LAD_MAY_2024_UK_BGC.shp")%>%
       st_transform(27700)  

lad <- UKlad |>
  rename(lad_code = LAD24CD,
         lad_name = LAD24NM) |>    # 方便后面 join 后清晰
  select(lad_code, lad_name, geometry)


CO2emission <- CO2emission %>%
  rename(lad_code = `Local Authority Code`,
         lad_name = `Local Authority`,
         pop_k    = `Mid-year Population (thousands)`,
         area_km2 = `Area (km2)`
  )


UKemission <- lad %>%                               
  left_join(CO2emission, by = c("lad_code", "lad_name")) %>%  
  filter(!if_any(ends_with("_pc"), is.na))             

plot(st_geometry(UKemission), col = "grey")  # 可视化所有区域
```

### 2.0 spatial weights and clustering
```{r}
library(sf);        # 读/写 sf 对象
library(dplyr);     # 数据整理
library(cluster);   # silhouette
library(dbscan);    # DBSCAN
library(tmap)       # 地图

tmap_mode("plot")   # 想交互改 "view"

sector_pc <- c("Agriculture_pc", "Industry_pc", "Commercial_pc",
               "Domestic_pc", "Public Sector_pc",
               "Transport_pc", "Waste_pc", "LULUCF_pc")

X <- UKemission |>
       st_drop_geometry() |>
       select(all_of(sector_pc)) |>
       mutate(across(everything(), ~ replace_na(.x, 0))) |>
       scale() |>                   # Z-score 标准化
       as.data.frame()


```

```{r}
sil_vals <- sapply(4:10, function(k){
  km <- kmeans(X, centers = k, nstart = 50)
  mean(silhouette(km$cluster, dist(X))[, 3])
})

best_k <- which.max(sil_vals) + 3   # 因为从 4 开始
cat("最佳 k =", best_k, "平均 silhouette=", round(sil_vals[best_k-3], 3), "\n")

set.seed(123)
km_final <- kmeans(X, centers = best_k, nstart = 50)
UKemission$kmeans_cl <- factor(km_final$cluster)

```

DBSCAN
```{r}
kNNdistplot(X, k =  8)             # 8 ≈ 2×维度，对应常用 minPts
abline(h = 0.25, col = "red")      # 0.25 只是示例；看拐点自定

db <- dbscan(X, eps = 0.25, minPts = 8)
table(db$cluster)                  # 0 表示噪声
UKemission$dbscan_cl <- factor(db$cluster)

```
画图
```{r}
table(UKemission$kmeans_cl)

centroids <- UKemission |>
  st_drop_geometry() |>
  group_by(kmeans_cl) |>
  summarise(across(all_of(sector_pc), mean)) |>
  arrange(kmeans_cl)

print(centroids, n = Inf)

write.csv(centroids,
          file = "results/kmeans_centroids.csv",
          row.names = FALSE)

```

```{r}

tm <- tm_shape(UKemission) +
        tm_fill("kmeans_cl",
                palette = "Set3",
                title   = "k-means (k = 5)") +
        tm_borders(col = "grey50", lwd = 0.2)

tmap_save(tm,
          filename = "results/kmeans_map.png",
          dpi       = 300,          # 分辨率：300 dpi 适合论文/幻灯片
          width     = 2000,         # 像素宽度
          height    = 3000)         # 像素高度

```






