---
title: "spatial_clustering"
output: html_document
date: "2025-07-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Spatial Clustering for UK 2022 GHG emission 


### 1.0 data cleaning-loading packages
```{r}
library(readr);    library(dplyr)
library(sf);       library(spdep)
library(FactoMineR);library(factoextra)
library(cluster);  library(tmap)
tmap_mode("plot")
```

### 1.1 reading GHG data for 2022
```{r}

ghg <- read_csv(
  "data/2005-2022-local-authority-ghg-emissions-csv-dataset.csv",
  show_col_types = FALSE
)

ghg22 <- ghg %>% filter(`Calendar Year` == 2022)
```

### 1.2 selecting out all sectors I need
```{r}
# select out all the sectors 
# Also I only need the GHG is CO2
library(dplyr)

target_sectors <- c(
  "Agriculture",
  "Industry",
  "Commercial",
  "Domestic",
  "LULUCF",
  "Public Sector",  
  "Transport",
  "Waste"
)

# filter out GHG == "CO2"
CO2 <- ghg22 %>%
  filter(
    `LA GHG Sector` %in% target_sectors,
    `Greenhouse gas` == "CO2"
  )

# summarise all CO2 territorial emission 
CO2_sum <- CO2 %>%
  group_by(`Local Authority Code`, `LA GHG Sector`) %>%
  summarise(kt_CO2 = sum(`Territorial emissions (kt CO2e)`), .groups = "drop")

# turn multi-variable into columns
library(tidyr)
CO2_sector <- CO2_sum %>%
  pivot_wider(names_from = `LA GHG Sector`,
              values_from = kt_CO2,
              values_fill  = 0)

CO2emission <- CO2_sector %>%
  left_join(select(ghg22, `Local Authority`,`Local Authority Code`, `Mid-year Population (thousands)`, `Area (km2)`) %>% distinct(),
            by = "Local Authority Code") %>%
  mutate(across(all_of(target_sectors),
                ~ .x / `Mid-year Population (thousands)`,  # t CO2 / person
                .names = "{.col}_p1000"))

#clean the df
CO2emission <- CO2emission %>%
  select(
    `Local Authority Code`,
    `Local Authority`,
    `Mid-year Population (thousands)`,
    `Area (km2)`,
    Agriculture,
    Agriculture_p1000,
    Industry,
    Industry_p1000,
    Commercial,
    Commercial_p1000,
    Domestic,
    Domestic_p1000,
    `Public Sector`,
    `Public Sector_p1000`,
    Transport,
    Transport_p1000,
    Waste,
    Waste_p1000,
    LULUCF,
    LULUCF_p1000
  )

```

p.s. Here territorial emissions covers all emissions happened in that geographical area
The subset table is CO2 emissions within the scope of influence of LAs, which Only CO₂ sources that local councils can realistically shape through local policy and services.

###1.3 Loading UK LAD map and merge it with CO2 emission
```{r}
# 1. read map
UKlad <- sf::st_read("data/LAD_boundaries/LAD_MAY_2024_UK_BGC.shp")%>%
       st_transform(27700)  

lad <- UKlad |>
  rename(lad_code = LAD24CD,
         lad_name = LAD24NM) |>    # 方便后面 join 后清晰
  select(lad_code, lad_name, geometry)


CO2emission <- CO2emission %>%
  rename(lad_code = `Local Authority Code`,
         lad_name = `Local Authority`,
         pop_k    = `Mid-year Population (thousands)`,
         area_km2 = `Area (km2)`
  )


UKemission <- lad %>%                               
  left_join(CO2emission, by = c("lad_code", "lad_name")) %>%  
  filter(!if_any(ends_with("_p1000"), is.na))    

sf::st_write(UKemission, "results/UKemission.gpkg", delete_dsn = TRUE)


```

### 2.0 spatial weights and clustering
```{r}
library(sf);        # 读/写 sf 对象
library(dplyr);     # 数据整理
library(cluster);   # silhouette
library(dbscan);    # DBSCAN
library(tmap)       # 地图

tmap_mode("plot")   # 想交互改 "view"

sector_p1000 <- c("Agriculture_p1000", "Industry_p1000", "Commercial_p1000",
               "Domestic_p1000", "Public Sector_p1000",
               "Transport_p1000", "Waste_p1000", "LULUCF_p1000")

X <- UKemission |>
       st_drop_geometry() |>
       select(all_of(sector_p1000)) |>
       mutate(across(everything(), ~ replace_na(.x, 0))) |>
       scale() |>                   # Z-score 标准化
       as.data.frame()



```

```{r}
set.seed(42)
sil_vals <- sapply(4:10, function(k){
  km <- kmeans(X, centers = k, nstart = 50)
  mean(silhouette(km$cluster, dist(X))[, 3])
})

best_k <- which.max(sil_vals) + 3   # 因为从 4 开始
cat("最佳 k =", best_k, "平均 silhouette=", round(sil_vals[best_k-3], 3), "\n")

set.seed(99)
km_final <- kmeans(X, centers = best_k, nstart = 50)
UKemission$kmeans_cl <- factor(km_final$cluster)

```

DBSCAN
```{r}
kNNdistplot(X, k =  16)             # 8 ≈ 2×维度，对应常用 minPts
abline(h = 3.5, col = "red")      # 0.25 只是示例；看拐点自定

db <- dbscan(X, eps = 3.5, minPts = 16)
table(db$cluster)                  # 0 表示噪声
UKemission$dbscan_cl <- factor(db$cluster)



```
### 1.3. kmean=5时，cluster表格
```{r}
table(UKemission$kmeans_cl)

centroids <- UKemission |>
  st_drop_geometry() |>
  group_by(kmeans_cl) |>
  summarise(
    n_LAD   = n(),                    # 先算数量
    across(all_of(sector_p1000), mean)
  ) |>
  arrange(kmeans_cl) |>
  mutate(across(all_of(sector_p1000), round, 2))




print(centroids, n = Inf)

write.csv(centroids,
          file = "results/kmeans_centroids.csv",
          row.names = FALSE) 

```


### 1.3.2 kmean=5时，cluster map
```{r}
UK_map <-
 tm_shape(UKemission) +
        tm_fill("kmeans_cl",
                palette = "Set3",
                title   = "k-means (k = 5)") +
        tm_borders(col = "grey50", lwd = 1)+
        tm_layout(legend.title.size = 1.5,   # 图例标题放大
            legend.text.size  = 2) 


tmap_save(UK_map, "results/UK_kmeans_print.pdf",
         width = 50, height = 50, units = "cm")

```
```{r}
london_bbox <- st_bbox(c(xmin = -0.55, xmax = 0.30,
                         ymin = 51.20, ymax = 51.80),
                       crs = 4326)

#—— ❷ 画图：裁剪 + 放大 + 调整图例字号                         ----
london_map <- 
tm_shape(UKemission, bbox = london_bbox) +       # ← 只显示伦敦
  tm_fill("kmeans_cl",
          palette = "Set3",
          title   = "k-means (k = 5)") +
  tm_borders(col = "grey50", lwd = 1) +
  tm_layout(legend.title.size = 1.4,   # 图例标题放大
            legend.text.size  = 1.2)   # 图例内容放大

tmap_save(london_map, "results/london_kmeans_print.pdf",
          width = 17, height = 22, units = "cm")
```

### 2.1 Running Silhouette and Gap, and draw PCA
```{r}
## ------------ 前期准备 ------------
library(sf)
library(cluster)      # silhouette()
library(factoextra)   # fviz_ 系列
library(FactoMineR)   # PCA()
library(ggplot2)
library(ggrepel)

set.seed(123)          # 可复现

# 假设 UKemission 里已经有 kmeans_cl 这个因子列
df  <- UKemission |>
  st_drop_geometry() |>
  mutate(across(all_of(sector_p1000), scale))        # 标准化用于 silhouette / PCA

## ----------------------------------
## 1) Silhouette plot
## ----------------------------------
## ---- 重新生成 silhouette 对象 ----
cl_vec    <- as.integer(df$kmeans_cl)              # ← 把 factor 变成 1,2,3...
dist_mat  <- dist(select(df, all_of(sector_p1000))) # 距离矩阵保持不动

sil       <- silhouette(cl_vec, dist_mat)

p_sil <- fviz_silhouette(sil) +
  ggtitle("Silhouette Plot for k-means (k = 5)")

ggsave("silhouette_k5.png", p_sil, width = 7, height = 5, dpi = 300)


## ----------------------------------
## 2) Gap Statistic
## ----------------------------------
gap <- clusGap(select(df, all_of(sector_p1000)),
               FUNcluster = kmeans,
               K.max = 10,  # 想测试到多少簇
               B = 300)     # bootstrap 次数

p_gap <- fviz_gap_stat(gap) + ggtitle("Gap Statistic")

ggsave("results/gap_statistic222.png", p_gap, width = 7, height = 5, dpi = 300)

## ----------------------------------
## ----------------------------------
## 3) PCA scatter, colored by cluster
##    并高亮簇 3 (n = 1) 和簇 5 (n = 4)
## ----------------------------------
# ① 做 PCA（无图）
pca_res <- PCA(select(df, all_of(sector_p1000)), graph = FALSE)

# ② 把前两主成分 + 聚类标签收进 tibble
pca_df <- as.data.frame(pca_res$ind$coord[, 1:2]) |>
  dplyr::rename(PC1 = Dim.1, PC2 = Dim.2) |>
  dplyr::mutate(
    LAD     = rownames(df),        # LAD 名
    cluster = df$kmeans_cl         # 原来 factor，就保留 factor
  )

# ③ 画散点 + 椭圆
library(ggplot2)
library(ggrepel)

p_pca <- ggplot(pca_df, aes(PC1, PC2, fill = cluster)) +
  geom_point(shape = 21, size = 2, alpha = 0.8, colour = "black") +
  stat_ellipse(aes(colour = cluster), type = "norm", level = 0.95, linetype = 2) +
  scale_fill_brewer(palette = "Set1") +
  scale_colour_brewer(palette = "Set1", guide = "none") +
  labs(
    title = "PCA of Sector Emissions (colored by cluster)",
    fill  = "Cluster"
  ) +
  theme_minimal(base_size = 13)

# ④ 给簇 3 & 5 的 LAD 打标签
p_pca <- p_pca +
  geom_text_repel(
    data = subset(pca_df, cluster %in% c(3, 5)),
    aes(label = LAD),
    size = 3,
    box.padding = 0.4,
    min.segment.length = 0
  )

ggsave("results/pca_clusters222.png", p_pca, width = 7, height = 5, dpi = 300)

out_id  <- which(df$kmeans_cl %in% c(1, 3, 5))   # 单点 & n=4 小簇
core_df <- df[-out_id, ]                          # 主干 300+ LAD

UKemission$lad_name[out_id]
```

### 2.2 After get rid of the outliers, checking the best k and re-cluster the results (core-df)
```{r}
## ---------------------------------------------------------
## 0. Which are the clusters, which are outliers？
## ---------------------------------------------------------
outlier_clusters <- c(1, 3, 5)       


## ---------------------------------------------------------
## 1.  core_df （get rid of outlier LAD）
## ---------------------------------------------------------
core_df <- UKemission |>
  filter(!(kmeans_cl %in% outlier_clusters)) |>
  st_drop_geometry()                  

# standardise
core_scaled <- core_df |>
  mutate(across(all_of(sector_p1000), scale))

## ---------------------------------------------------------
## 2. select k, (k=2 to k=4)
## ---------------------------------------------------------
library(factoextra)

set.seed(123)
fviz_nbclust(select(core_scaled, all_of(sector_p1000)),
             kmeans, method = "silhouette",
             k.max = 6, nstart = 30) +
  ggtitle("Silhouette method on core_df")


gap_core <- clusGap(select(core_scaled, all_of(sector_p1000)),
                    FUNcluster = kmeans, K.max = 6, B = 300)
fviz_gap_stat(gap_core) + ggtitle("Gap Statistic on core_df")

## ---------------------------------------------------------
## 3. 跑 k-means（先 k = 2，再视情况 k = 3）
## ---------------------------------------------------------
set.seed(123)
km_core2 <- kmeans(select(core_scaled, all_of(sector_p1000)), centers = 2, nstart = 30)
km_core3 <- kmeans(select(core_scaled, all_of(sector_p1000)), centers = 3, nstart = 30)

## ---------------------------------------------------------
## 4. return the result to core_df
## ---------------------------------------------------------
core_df$k2 <- factor(km_core2$cluster)
core_df$k3 <- factor(km_core3$cluster)

## ---------------------------------------------------------
## 5. Visualise
## 5a. silhouette for k = 2 / 3
## ---------------------------------------------------------
library(cluster)
sil_k2 <- silhouette(as.integer(core_df$k2),
                     dist(select(core_scaled, all_of(sector_p1000))))
sil_k3 <- silhouette(as.integer(core_df$k3),
                     dist(select(core_scaled, all_of(sector_p1000))))

fviz_silhouette(sil_k2) + ggtitle("Silhouette (k = 2, core_df)")
fviz_silhouette(sil_k3) + ggtitle("Silhouette (k = 3, core_df)")

## 5b. PCA scatter（再上色）
pca_core <- PCA(select(core_scaled, all_of(sector_p1000)), graph = FALSE)
pca_df   <- as.data.frame(pca_core$ind$coord[, 1:2]) |>
  rename(PC1 = Dim.1, PC2 = Dim.2) |>
  mutate(k2 = core_df$k2,
         k3 = core_df$k3)

ggplot(pca_df, aes(PC1, PC2, fill = k2)) +
  geom_point(shape = 21, size = 2, colour = "black") +
  stat_ellipse(aes(colour = k2), type = "norm", level = 0.95, linetype = 2) +
  scale_fill_brewer(palette = "Set1") +
  scale_colour_brewer(palette = "Set1", guide = "none") +
  theme_minimal(base_size = 13) +
  labs(title = "PCA on core_df (k = 2)")

## ---------------------------------------------------------
## 6. looking at centroids at each clusters and number of LADs
## ---------------------------------------------------------
library(dplyr)

centroids_k2 <- core_df |>
  group_by(k2) |>
  summarise(
    n_LAD = n(),
    across(all_of(sector_p1000), mean, .names = "{.col}_mean")
  )

centroids_k3 <- core_df |>
  group_by(k3) |>
  summarise(
    n_LAD = n(),
    across(all_of(sector_p1000), mean, .names = "{.col}_mean")
  )

print(centroids_k2, n = Inf)
print(centroids_k3, n = Inf)

```




